---
title: "Assignment 2 -- Mean-Variance Optimization Hands-on Exercise"
output:
  pdf_document:
    toc: yes
  html_document:
    code_folding: show
    number_sections: yes
    toc: yes
    toc_float: yes
---

# Final Deliverable

You may use any programming language that you prefer. In your final submission, answer the following questions:

1. How many rows of data do you delete in the price matrix? How many rows does the returns matrix have? (5pt).
2. Is the covariance matrix for the 50 selected stocks positive definite? (5pt)
3. What are the mean return and variance for the minimum risk portfolio constructed from the 50 selected risky assets? (10pt)
4. Suppose you have \$1,000 and would like to invest in the minimum risk portfolio, how much do you invest in IBM? (5pt)
5. If your risk-tolerance parameter is 0.1, what are the mean return and variance of the parametric efficient portfolio that's suitable for you? (10pt)
6. Suppose you have \$1,000 and would like to invest in the parametric-efficient portfolio, how much do you invest in IBM? (5pt)
7. Plot an efficient frontier plot in the $(\sigma,\mu)$-plane along with the 50 individual assets. (10pt)

# Summary of practical problems

In this exercise, we will see the following practical problems in financial analysis:
* Missing data: Two of the stocks in the dataset has some missing data. This is troublesome in many ways (for example, cannot estimate covariance matrix).
- A simplistic solution is to truncate the dataset to a narrower observation window where all stocks have valid prices.
* Estimated covariance matrix is NOT positive definite, which may seem strange at first. There are many ways this can happen. In our case, this is because we have more stocks (about 500) than available data (about 200).
- In practice, you may want to find ways to collect more data.
- In this assignment, we consider mean-variance portfolio optimization for a small subset of the stocks.

# Clean up workspace and load data

The following codes show how you can clear temporary variables (e.g., some variables left from previous codes) and load data from the .csv file provided.

2. Calculate the daily return from the daily prices. **You need to complete this step**. When you are done:
- The first 3 days' returns for "Stock A" are -0.016055947, 0.002956187, and 0.003065364.
- There are 253 stocks prices, so you should only have 252 returns. 

```{r}
rm(list=ls())     # clear temporary variables
library(ggplot2)  # library for plotting

prices <- read.csv("HistoricalPrices.csv",header = T)    # read data from file
prices[1:10,1:10] # visualize the historical price data
```

# Preliminary data manipulation (missing data particularly)
Seeing the first first columns and rows of the "price" matrix, we determine that the first two columns are irrelevant so we can remove them:

Suppose $A$ is a matrix in R, then $A[i,j]$ calls the $(i,j)$-th element of the matrix. Also, $A[i,]$ and $A[,j]$ calls the $i$-th row and the $j$-th column of the matrix. The minus ('-') sign in "prices[,-c(1,2)]" tells R to remove Column 1 & Column 2 of the "prices" matrix.

Your task:
1. Take a look at the columns of the "prices" matrix with titles "CARR" and "OTIS", you should see some NAs, i.e., missing data.
- This is because CARR and OTIS joined S&P 500 in March 2020. So these two stocks' price data is available only March 20.
2. Remove just enough rows of the "prices" matrix so that all the NAs in these two stocks are removed.
- By removing the rows, you will be removing some stock prices for other stocks too.
- Such removal makes sure that all the stocks have the same amount of data for later analysis. This is a simplistic solution to dealing with missing data.
- When you are done, the first stock price for "CARR" is $11.89012.
3. Calculate the returns matrix based on the resulting prices matrix.
- When you are done, the first three returns for "Stock A" are, 0.01064475, -0.04649420, and 0.05728266.
- The returns matrix should have 1 less row than

```{r}
prices <- prices[,-c(1,2)]    # remove the first two irrelevant columns 

# Your codes may be longer than 1 line, but please use the variable name "prices".
prices <- #FILL YOUR CODES HERE#    # remove some rows so all stock prices are available

# complete the following line
# Your codes may be longer than 1 line, but please use the variable name "returns".
returns <- #FILL YOUR CODES HERE#   # calculate returns from prices
```

# Stock returns and risks
There is nothing you need to complete in this section. This section is simply a re-run of our synchronous time activity on Jan 18, 2021. This time, the missing data is removed.

Indeed we see a positive slop, i.e., risk-reward trade-off. But again, what we see highly depends on the observation window and the observation frequency of the stock data that we use.
```{r}
# mean return and standard deviation
mean.return <- colMeans(returns)
std.return <- apply(returns, 2, sd)
summary(lm(mean.return ~ std.return))

# Organize data for plotting
Data <- data.frame(return = mean.return, risk = std.return)

# Plot scatter plot of stock returns vs. risks & line of best fit
p1 <- ggplot(Data, aes(y=return, x=risk)) +
  geom_point(alpha = .15, size=2, col= "darkorange") +
  geom_smooth(formula = y~x,method = "lm", se=F, lwd=1.5, col="#67a9cf") +
  ggtitle("Linear regression model fit")

p1
```

# Mean-variance Optimization -- First Attempt
There is nothing you need to complete in this section.

Please pay attention to how the mean return vector and the covariance matrix are estimated.
A new library is loaded to check if the estimated covariance matrix is positive definite, which is an assumption required by mean-variance optimization.

The answer turns out to be NO?! 

* One reason for non-positive-definite covariance matrix is that we have more stocks (about 500) than data (about 200). 
* One consequence of non-positive-definite covariance matrix is that it is not invertible. 
* We need the inverse of the covariance matrix to calculate the optimal portfolios, as we will see later.

```{r, error=TRUE}
mu.vec <- matrix(mean.return, ncol = 1) # mean vector, make sure it is a column vector
Sigma.mat <- cov(returns)               # covariance matrix

library(matrixcalc)   # a library to check if a matrix is positive definite
is.positive.definite(Sigma.mat) # the covariance matrix fails the positive definite test

Sig.inv <- solve(Sigma.mat) # invert the covariance matrix, or report error if not possible
```

# Mean-variance Optimization: Preliminary Analysis
We will next consider a mean-variance optimization problem for 50 stocks instead. With less stocks to consider, the estimated convariance matrix is now positive definite. As a result, we can calculate the optimal portfolios and visualize the efficient frontier.

Your tasks:

1. Calculate the returns of the 50 stocks specified in StockNames.csv.
2. Estimate the returns and standard deviations of the 50 stocks specified in StockNames.csv.
3. Plot the risk vs. return for the 50 stocks and the resulting line-of-best-fit. *This step is done for you if you keep the variable names as suggested.*
4. Calculate the mean vector and covariance matrix for the 50 stocks. Then verify if the covariance matrix is positive definite.
```{r}
stocks <- read.csv("DATADIRECTORY/StockNames.csv")[,2]  # the second column contains stock names

# Your codes may be longer than 1 line, but please keep the suggested variable names
returns <- #FILL YOUR CODES HERE#   # calculate the returns for the selected 50 stocks
mean.return <- #FILL YOUR CODES HERE#   # calculate the mean returns for the selected 50 stocks
std.return <- #FILL YOUR CODES HERE#   # calculate the standard deviation for the selected 50 stocks

Data <- data.frame(return = mean.return, risk = std.return)

ggplot(Data, aes(y=return, x=risk)) +
    geom_point(alpha = 0.5, size=2, col= "darkorange") +
    geom_smooth(formula = y~x,method = "lm", se=F, lwd=1.5, col="#67a9cf") +
    ggtitle("Linear regression model fit")

mu.vec <- #FILL YOUR CODES HERE#         # mean vector, make sure it is a column vector
Sigma.mat <- #FILL YOUR CODES HERE#      # covariance matrix

is.positive.definite(Sigma.mat) # the covariance matrix fails the positive definite test
```

As we can see, the covariance matrix passes the positive definiteness test.

# Mean-Variance Optimization: Optimal Portfolios
In this Section, you will be asked to use the formulas we derived in class to performn mean-variance portfolio optimization and answer a few questions based on your calculations.

Your tasks:

1. Calculate the minimum risk portfolio as well as its mean return and standard deviation.
2. Based on your calculations, answer Questions 3 & 4 in the "Final Deliverable" section.
3. Calculate the "zero-covariance" portfolio as well as its mean return and standard deviation.
4. Based on your calculations, answer Questions 5 & 6 in the "Final Deliverable" section.

```{r}
Sigma.inv <- solve(Sigma.mat)     # calculate the inverse of the covariance matrix  
ell.vec <- matrix(1,ncol = 1, nrow = 50) # define a column vector of 1's

Sigma.ell <- Sigma.inv %*% ell.vec                    # maxtrix-vector product
ell.Sigma.ell <- as.numeric(t(ell.vec) %*% Sigma.ell) # inner product of vectors 
# The "as.numeric" may seems strange. But in R, dividing an 1-by-1 matrix is an 
# invalid calculation. So we need to translate the 1-by-1 matrix into a number first.

################################
# complete the following lines #
################################
x.m <- #FILL YOUR CODES HERE#            # min-risk portfolio
mu.m <- #FILL YOUR CODES HERE#       # mean return of min-risk portfolio
sigma2.m <- #FILL YOUR CODES HERE#  # variance of min-risk portfolio
mu.m
sigma2.m
x.m["IBM",]

Sigma.mu <- #FILL YOUR CODES HERE# 
ell.Sigma.mu <- #FILL YOUR CODES HERE# 

x.z <- #FILL YOUR CODES HERE#            # zero-covariance portfolio
mu.z <- #FILL YOUR CODES HERE#       # mean return of zero-covariance portfolio
sigma2.z <- #FILL YOUR CODES HERE#  # variance of zero-covariance portfolio

tau <- 0.1
x.tau <- #FILL YOUR CODES HERE#            # parametric-effieicnt portfolio
mu.tau <- #FILL YOUR CODES HERE#       # mean return of parametric-effieicnt portfolio
sigma2.tau <- #FILL YOUR CODES HERE#  # variance of parametric-effieicnt portfolio
mu.tau
sigma2.tau
x.tau["IBM",]
```

* A3: The mean return and variance of the min-risk portfolio are 0.0748% and $7.23\times 10^{-5}$, respectively.
* A4: With $1,000 initial investment, I will **short-sell** about $65.49 in IBM.
* A5: The mean return and variance of the parameteric-efficient portfolio for $\tau=0.1$ are 1.40% and $0.0014$, respectively.
* A6: With $1,000 initial investment, I will **short-sell** about $861 in IBM.

# Mean-Variance Optimization: Efficient Frontier
In this Section, you will be asked to use the optimal portfolios calculated above to plot the efficient frontier.

Your tasks:

1. Based on the set of risk-tolerance parameters provided, calculate the corresponding parametric efficient portfolios' mean returns and variances.
    - Hint: You can, but do not have to calculate the optimal portfolios first. 
2. Plot the efficient frontier in the $(\sigma,\mu)$-plane along with the 50 individual assets. Also indicate the parametric-efficient portfolio for $\tau=0.1$ in the plot.

```{r}
tau <- seq(0, 0.15, length.out = 100)

################################
# complete the following lines #
################################
mu.p <- #FILL YOUR CODES HERE#        # mean returns of parametric-effieicnt portfolios
sigma2.p <- #FILL YOUR CODES HERE#    # variances of parametric-effieicnt portfolio

Data2 <- data.frame(mu = mu.p, sigma = sqrt(sigma2.p))
Data3 <- data.frame(mu = mu.tau, sigma = sqrt(sigma2.tau))

ggplot(Data, aes(y=return, x=risk)) +
    geom_point(alpha = 0.5, size=2, col= "darkorange") +
    geom_line(data = Data2, aes(x=sigma, y=mu), color = "#67a9cf", lwd = 1.5) +
    geom_point(data = Data3, aes(x=sigma, y=mu), color = "red", size = 5, shape=13) +
    ggtitle("Individual assets and risky efficient frontier")
```











